---
title: "The Presidency"
author: "Josh Allen"
date: '`r Sys.Date()`'
output:
  xaringan::moon_reader:
    css: ["default"]
    self_containted: true
    nature:
      slideNumberFormat: "%current%"
      highlightStyle: github
      highlightLines: true
      ratio: 16:9
      countIncrementalSlides: false 
      navigation:
        scroll: true
---

```{r preamble, child = here::here('preamble.Rmd')}
```



---
# Looming Deadlines

.bold[Misinfo Activity]


.bold[Core Assessment]: April 25th


???
The Core assessment is critical because it is a university requirement. So you have to do it.

TBH the pollen is kind of 
---

---

## Qualifications

- 35 years old

- Natural born citizen

- Retain residency in the US for at least 14 years

- 12th Amendment (1804) 

- 22nd Amendment (1951) fixes the number of terms to 2

???
These are the constitutional qualifications to be president as set aside in Article II, Section I, Clause V.  (the twelfth amendment also extends these qualifications to the vice-president (1804).) 

Natural born – must be a born in the United States – differentiation from individuals that attain citizenship 


The 12th amendment basically just extends the minimum qualifications for president to anybody in the line of succession


22nd amendment – the Constitutional amendment that limits the number of terms that a president can serve to two. 

---

---
## Constitutional Sources of Power


.pull-left[

### Expressed Powers 

- Derived from Article II USC
- Head of the Executive Branch (bureaucracy)
- Commander-in-chief
- Enter in treaties with foreign nations
- Make judicial/executive appointments



]




.pull-right[
### Implied Powers

- Derived from the “Take Care Clause” of Article II USC
- Organizing the Federal bureaucracy (create executive - departments: Homeland Security)
- Issue Executive Orders


]

???
Enumerated

Military: President is the highest military authority in the USA, with control of the entire military establishment. Also heads all intelligence agencies. 

This point can spark some discussion about how the president has used his power as commander in chief to expand his power, but it’s covered more later in the presentation. 


No President that has served has served in the Marines or in the Coast Guard. We have had 3 presidents  that have been Generals in the Army (Washington, Grant, and Eisenhower). We also have only had 12 presidents that have no military service.  15 presidents served in the Reserves and 9 served in State militia. 


Judicial
 Nixon. Historical link to how the President can be one where supplicants can seek mercy, like Kings used to do. 

Nixon – pardoned for crimes by Ford after his resignation from the presidency for abuses of power. Ford pardon came as an effort more for public opinion and resurgence in faith in the government rather then because he wasn’t guilty. 

Turkey pardons – By far the most famous of the all pardons is the one that occurs each year at Thanksgiving of the presidential turkey. This has been a tradition at the White House since H.W. Bush was in office (1989). Generally, the turkeys are sent to petting zoo’s across the nation.  It is a national contest each year that is run by the white house, last years turkeys were named Popcorn and Caramel. Unfortunately, Popcorn won the popular vote, but eventually both Turkeys were spared.  Unfortunately, the common trend of obesity in these turkeys eventually cause them to die within a year of being pardoned.) 

Diplomatic – rise of Executive Orders over Legislative law due to difficulty of securing the consent of the senate.  Executive orders have been on the rise since the political division between the President and Congress has grown more volatile. 
Executive – talk about appointments of Judges and heads of Agencies. 

Executing all laws, too, shows that the president is bound to execute laws that he/she may not agree with. 
Advice and Consent – the president seeks advice from individuals (Senatorial Curtesy) on whom to nominate for federal positions. After vetting, the Senate votes on the appointment


Implied Powers

Legislative - Executive orders are a regulation issued by the president that has the effect and formal status of legislation. Explain Veto and Pocket veto. 

The Roosevelt Presidents issued the most number of presidential pardons during their presidencies ( FDR – 3500 and Teddy – 1281). Coolidge and Wilson follow in close third and fourth with over 1000 each. 

Executive Orders – are orders signed into effect by presidents in order to ensure that the government and all of its agencies can essentially function. They have the full force of law

---

---
## The Seven Roles of the Presidency


- Executive
- Legislative
- Judicial
- Diplomatic
- Military
- Public Opinion
- Political Party




---

---
## President as Executive Leader

- Executes the law
- Appoints Department Heads
- Recommends federal budget
- Issues Executive Orders
???

Executing all laws, too, shows that the president is bound to execute laws that he/she may not agree with. 
Advice and Consent – the president seeks advice from individuals (Senatorial Curtesy) on whom to nominate for federal positions. After vetting, the Senate votes on the appointment
---


---
## Executive Orders 

- Executive Orders are NOT laws written by the President

- Executive Orders are declarations issued by a president - that relate to an organization of the federal bureaucracy

 Examples:
- President Nixon uses EO to establish EPA (Regulatory - bureaucratic agency)

- President Eisenhower uses EO to desegregate schools

- President Obama issues EO to keep notice of criminal record of federal job applications (11/02/2015)

???
 Executive orders are issued by the President of the United States, acting in his capacity as head of the executive branch, directing a federal official or administrative agency to engage in a course of action or refrain from a course of action. They are enforceable to the extent that the represent a valid exercise of the President’s power (i.e. the action must be within the president’s constitutional authority).



---
---

```{r}
vembedr::embed_url("https://www.youtube.com/watch?v=oyOf3g-PJ94")
```


---

---
## President as Legislative Leader

- Makes policy recommendations

- Signs/Vetoes legislation

- Delivers the State of the Union Address

- Convene/Adjourns Congress
???
The last bullet point is technically true but has not really been used. There are also some parliamentary tricks to keep congress in session because the president can make appointments to an office while they are out of session that are automatically confirmed. This is called a recess appointment 


In modern times, the Senate is in session nearly year-round, making the recess appointment mechanism far less necessary or useful for upkeep of government function. Nonetheless, in recent times this power has also been controversially used as a political tool to temporarily install an unpopular nominee by sidestepping the Senate's role in the confirmation process
---

---
## Presidents as Judicial Appointers

- Appoints justices to the Supreme Court
- Appoints judges to federal courts
- Grants pardons

???
Although the president’s power is broad, it is not without accepted limitations. Perhaps the most important is that the president can only pardon federal offenses; he cannot interfere with state prosecutions. Also, the pardoning power only extends to criminal offenses; it does not preclude civil actions. 

Unlike a commutation, which shortens or eliminates an individual’s punishment, a pardon absolves the individual of guilt. For example, President Trump commuted Roger Stone’s prison sentence so that Mr. Stone did not serve the punishment for his guilty conduct. 

---


---
## Chief Diplomat 

- Appoints ambassadors
- Enters treaties with foreign countries
- Enters into executive agreements
- Serves as head of state

???
Diplomatic – rise of Executive Orders over Legislative law due to difficulty of securing the consent of the senate.  Executive orders have been on the rise since the political division between the President and Congress has grown more volatile. 
Executive – talk about appointments of Judges and heads of Agencies. 
---

---
## Commander in Chief

- Serves as commander-in-chief over the US military
- Presidents’ interpretation of “commander in chief” has allowed the deployment of troops without Congressional - authorization
- War Powers Resolution 1973

???
Following the Executive branches continued willingness to engage in military conflicts without a declaration of war, Congress passed the War Powers Resolution of 1973. This act requires the president to notify Congress within 48 hours of committing armed forces to military action, and requires the president to seek continued approval for troop deployments within the first 60 days

---


---

## Presidents as Party Leaders
- Implement party priorities

- Shape the party agenda

- Symbolic head of the political party



---




---
## Rise of Presidential Power



- History 
  - Washington – The Autonomy of the Executive 
  - Jackson – The man and his veto 
  - Lincoln – War Powers expansion 
  - T. Roosevelt – “The Bully” 
  - The “Communicators” – FDR, JFK, and Reagan 

- War Powers 

- Executive Privilege – the implied power of the president  and members of the executive to resist or deny subpoenas and demands of other branches of government to access information 


???
History – 
Executive Privilege and the autonomy of the Executive – the Founders never intended the President to be a very powerful body, but over time that role has developed and changed. GW wasted no time in declaring the autonomy of the executive to demands by Congress. 

Jackson – First president to ever used the presidential veto extensively. The first six presidents (1789-1829) used the veto only 9 times. Jackson in his two terms used it on 12 separate occasions. 

Lincoln – during the Civil War not only declared martial law ( extreme govt and military control placed upon territories/lands when in crisis, upheaval, or chaos). Also suspended the writ of habeas corpus which is the right of prisoners to be able to seek relief from what they believe to be false imprisonment. 

T-Roosevelt – the use of the Bully Pulpit to badger individuals/media into supporting his actions/wants in the legislature. 

The Communicators – these three presidents were known as the communicator presidents since they dominated the used of Media at the time to get their messages across to the American Public. This helped to sway many a politician towards legislation and action that would have been untold of before. 

War Powers – slowly but surely this power has grown over time. The President is a less deliberative or more reactionary body of the government so as a result sudden attacks or necessary military actions come under his purview. The War Powers resolution of 1973 allows for the President to commit troops anywhere for up to 90 days. Essentially expansion of this power have made Congress’s power to declare war obsolete.  It has only happened 5 times officially in history War of 1812, Mexican-American War, Spanish –American War, WWI, and WWII. Other situations have required Congress to appropriate money without official declarations of war. These were termed as extended military excursions or engagements including the Persian Gulf War, The Iraq War, and Vietnam War. Others have been funded by Congress after UN Security Council Resolutions for intervention have been issued including the Korean War and the recent civil war in Libya. 

One possible explanation is that Congress doesn’t want the responsibility of telling troops to come home after they’ve been deployed for 90 days (War powers act) and it’s better politically to support the president in times of conflict. 

Executive Privilege – Nixon and Watergate Scandal 

---

---
## Bully Pulpit and Policy



.pull-left[


```{r}
vembedr::embed_url("https://www.youtube.com/watch?v=Q3YhruUmPAM")
```

]

.pull-right[
- A platform through which a President can  push his agenda by swaying public opinion and the media and effect change.

- New Media and the President  ]


???
The term was coined by Teddy Roosevelt during his terms in office. He saw it as a good forum for advocating his agenda. He used it not only as a “bully” to get what he wanted, but also as a pat on the back to himself with the more commonly used term (“bully for you”). 
Some examples of the Bully Pulpit: 

FDR and “the switch in time that saved nine” 

Obama and the Affordable Care Act

It turns out that FDR’s landslide 1936 re-election was probably sufficient to swing CJ Hughes & Roberts. Roberts wrote a letter to Hughes in Dec ‘36 saying it was time to switch. 
Turns out the court-packing threat may have weakened FDR with Congress and the Court.


The relationship of the president and the media has developed significantly over time, giving the president vast amount of opportunities with which to utilize this power. This has extended well beyond traditional forms of media like the radio to blogs, twitter, and even Instagram. 





---


---
## State of the Union

.pull-left[

Article II, Section 3: the President “shall from time to time give to the Congress Information of the State of the Union, and recommend to their Consideration such Measures as he shall judge necessary and expedient.” ]

.pull-right[

```{r}
vembedr::embed_url("https://www.youtube.com/watch?v=0TKxu-oxue8")
```

]




???

Mandated by the Constitution, lately has been more of a show than meaningful. 
State of the Union is an important event in that it signals what direction the President would like to take legislation for that year that have been “agreed upon by the leadership of each of the parties”. (bipartisanship) 

It does have criticism, it is considered a mostly futile event in many cases since often the two parties do not stick to any of the proposed ideas during the following legislative term. 
The SOU has under gone many historical changes since the beginning of the presidency. 
Audience – 33.3 million viewers of the 2014 SOU on 13 different networks with worldwide coverage. This figure is lowest audience since SOU in 2000. 

Written speech transformed into oral speech – Washington gave first orally, but it wasn’t till Wilson (1913) that it changed back to an oral speech.  
Jimmy Carter was the last to give a written report in 1981. 

Media – First Radio – Calvin Coolidge (1923) 

First Recorded TV – Truman (1947) First Primetime Live Televised – L

BJ (1965) 

First live webcast SOU – Bush (2002) 

“State of the Union” term coined by FDR in 1934, but not popularized until 1947. 

 

---

---
## Structure of the Presidential Office

- White House Staff
- The Chief of Staff
- Executive Office of the President (EOP)
  - Established in 1933 by FDR 
  - National Security Council
  - President, VP, Secretaries of State, Defense, Treasury, - and Energy, NSA, Attorney General, and Joint Chiefs
  - Office of Management and Budget
  - Largest office 
- Vice President 
- Council of Economic Advisors 

???
The executive office of the president consists of the immediate staff of the president and 4 levels of support staff. They are headed by the White House Chief of Staff, and the immediate “chain of command” is housed in the West Wing of the White House. The rest of the support staff are housed in various support buildings with local access to the White House or Capitol Hill. 

Roles of the Chief of Staff

The responsibilities of the chief of staff are both managerial and advisory and may include the following:

Selecting senior White House staffers and supervising their offices' activities;

Managing and designing the overall structure of the White House staff system;

Control the flow of people into the Oval Office;

Manage the flow of information to and decisions from the Resolute Desk (with the White House staff secretary);

Directing, managing and overseeing all policy development

Protecting the political interests of the president;

Negotiating legislation and appropriating funds with United States Congress leaders, Cabinet secretaries, and extra-governmental political groups to implement the president's agenda; and

Advise on any and usually various issues set by the president.



EOP helps the president with various tasks, the Office of Management and Budget for example is responsible for coming up with a national budget, designing the president’s program, reporting on agency activities, and overseeing regulatory proposals.  The staff has grown exponential with the staff currently consisting of 2000 persons with an estimated budget of 375 million each year. 

Council of Economic Advisors constantly analyzes the economy and economic trends in order to give advice. 

National Security Council meets regularly to give advice on large national security issues

---

---
## The Cabinet

- 15 different member agencies 

- 8 additional cabinet officials
  - Chief of Staff 
  - US Trade Rep 
  - Director of Nat Intelligence 
  - UN Ambassador 
  - Director of Management and Budget 
  - CIA Director 
  - EPA 
  - Small Business Association

???
These individuals are the most senior appointed officials in the executive branch and the heads of the largest executive departments. These agencies make up fifteen of the “most” important federal agencies and meet more regularly with the President than heads of smaller agencies. 

These leaders are also appointed by the president, confirmed by the senate, and can be removed by the president.  

---
